---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-", 
  out.width = "100%"
)
library(ggplot2)
set.seed(1542)
```

[![Travis-CI Build Status](https://travis-ci.org/malcolmbarrett/ggdag.svg?branch=master)](https://travis-ci.org/malcolmbarrett/ggdag) [![AppVeyor Build status](https://ci.appveyor.com/api/projects/status/kd3ed7rj6p2vd36t?svg=true)](https://ci.appveyor.com/project/malcolmbarrett/ggdag)


# ggdag: An R Package for visualizing and analyzing directed acyclic graphs

Tidy, analyze, and plot directed acyclic graphs (DAGs). `ggdag` uses the powerful `dagitty` package to create and analyze structural causal models and plot them using `ggplot2` and `ggraph` in a consistent and easy manner. 

## Installation

You can install ggdag from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("malcolmbarrett/ggdag")
```

## Example

`ggdag` makes it easy to use `dagitty` in the context of the tidyverse. You can directly tidy `dagitty` objects or use convenience functions to create DAGs using a more R-like syntax:

```{r tidydag, dpi=300, message=FALSE}
library(ggdag)

#  example from the dagitty package
dag <- dagitty::dagitty( "dag {
    y <- x <- z1 <- v -> z2 -> y
               z1 <- w1 <-> w2 -> z2
               x <- w1 -> y
               x <- w2 -> y
               x [exposure]
               y [outcome]
               }")

tidy_dag <- tidy_dagitty(dag)

tidy_dag 

#  using more R-like syntax to create the same DAG
tidy_ggdag <- dagify(y ~ x + z2 + w2 + w1,
             x ~ z1 + w1,
             z1 ~ w1 + v,
             z2 ~ w2 + v,
             w1 ~~ w2, # bidirected path
             exposure = "x",
             outcome = "y") %>% tidy_dagitty()

tidy_ggdag
```

`ggdag` also provides functionality for analyzing DAGs and plotting them in `ggplot2`:

```{r ggdag, dpi=300}
ggdag(tidy_ggdag)
ggdag_adjustment_set(tidy_ggdag)
```

As well as geoms and other functions for plotting them directly in `ggplot2`:

```{r ggdag_geoms, dpi=300}
dagify(m ~ x + y) %>% 
  tidy_dagitty() %>% 
  node_dconnected("x", "y", controlling_for = "m") %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend, shape = adjusted, col = d_relationship)) +
    geom_dag_edges(aes(end_cap = ggraph::circle(10, "mm"))) +
    geom_dag_collider_edges() +
    geom_dag_node() +
    geom_dag_text(col = "white") +
    theme_dag() + 
    scale_dag(expand_y = expand_scale(c(0.2, 0.2)))
```

And common structures of bias:
```{r ggdag_common, dpi=300}
ggdag_equivalent_dags(confounder_triangle())

ggdag_butterfly_bias(edge_type = "diagonal")
```
